Ecosystem size and complexity dictate riverine biodiversity

### 2021-01-29 manuscript.Rmd
Update manuscript.Rmd

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  word_document: default
  html_notebook: default
abstract:
  The species-area relationship (the SAR), a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and formed the basis of area-based conservation strategies. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity), and its role in scaling biodiversity is largely unknown. Here, we hypothesized that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity (area x habitat heterogeneity), two factors that regulate watershed-scale species richness. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two biogeographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the conflicts between biodiversity conservation and societal demands.

@@

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@rinaldoEvolutionSelectionRiver2014; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grantLivingBranchesPopulation2007; @rinaldoRiverNetworksEcological2018;@tonkinRoleDispersalRiver2018]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure[@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @teruiMetapopulationStabilityBranching2018], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@altermattDiversityRiverineMetacommunities2013; @carraraComplexInteractionDendritic2014; @carraraDendriticConnectivityControls2012; @teruiMetapopulationStabilityBranching2018]. However, most research has explored the consequences of branching complexity for local community structure[@altermattRiverNetworkProperties2013; @brownDendriticNetworkStructure2010] or has relied solely on theoretical arguments with limited replications of branching architecture[@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where the accelerated species loss threatens the delivery of ecosystem services[@dudgeonFreshwaterBiodiversityImportance2006].

@@

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches where local communities inhabit[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a] (**Figure 1**). Two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see **Methods** for model details).

Given the results of extensive sensitivity analysis (**Supplementary Information**), we considered `r n_scenario` simulation scenarios, which comprised a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries [@chezikRiverNetworksDampen2017; @mooreEmergentStabilityLarge2015; @petersonModellingDendriticEcological2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities[@mooreSynchronizationPortfolioPerformance2010; @poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see **Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Information**). Under each simulation scenario, we simulated 1000 time steps of metacommunity dynamics in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

Our theoretical analysis yielded results consistent with our prediction. Ecosystem size and complexity both increased $\gamma$ diversity under a realistic landscape scenario (**Figures 2 and 3**), where the environmental variation at headwaters ($\sigma_h$) is greater than the degree of local environmental noise ($\sigma_l$) [@petersonModellingDendriticEcological2013; @teruiMetapopulationStabilityBranching2018]. The relationships had a characteristic of power-law (i.e., linear in a log-log scale) and were consistent under various ecological scenarios. The strength of competition and dispersal processes (dispersal distance and probability) did not change the relationships between $\gamma$ diversity and ecosystem properties (compare panels in **Figures 2 and 3**). Hence, the ecosystem size and complexity are expected to increase $\gamma$ diversity regardless of ecological scenarios.

@@

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data across two geographic regions: the Hokkaido island in Japan and the midwestern US (Midwest). These regions are located in comparable latitude ranges but support distinct fish communities (**Figure 1**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Methods** and **Supplementary Information**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area), or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). We combined this data set with geospatial information (watershed area, branching probability, annual mean temperature, annual cumulative precipitation, and the fraction of forest) to examine potential influences of macro-scale drivers of species diversity. Using this data set, we developed global and region-specific models for each of the diversity measures ($\alpha$, $\beta$, and $\gamma$ diversity) to examine whether observed patterns are consistent across the two geographic regions. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity by including interaction terms with the dummy binary variable of the region (Hokkaido = 0; Midwest = 1). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (see **Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

@@

The emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation[@poffHomogenizationRegionalRiver2007], landscape modification[@finlayStreamSizeHuman2011; @moritaEffectsHabitatFragmentation2002], and stream burial[@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research[@koningNetworkGrassrootsReserves2020; @stollSmallImpoverishedRegional2013; @sundermannRiverRestorationSuccess2011]. While large reserves or restoration sites are undoubtedly important[@koningNetworkGrassrootsReserves2020], our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin[@koningNetworkGrassrootsReserves2020]. Hence, our findings offer the ecological foundation for the complexity-based conservation strategies in rivers.

While the prevailing evidence supports the importance of ecosystem size in scaling species diversity, ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial[@turnerLandscapeEcologyTheory2015] and aquatic ecosystems[@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of ecosystem structure. Our general framework of the dual dimensional approach, ecosystem size and complexity, should apply beyond branching ecosystems and broaden viable options towards successful conservation. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

We depicted branching ecosystems as a bifurcating spatial network of connected habitat patches[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a] (**Figure 1**). Habitat patches, or local communities, can be either non-branching or branching river sections with a unit length $l$, which defines the scale of local species interactions. Two parameters determine the geometric properties of simulated branching networks: the number of habitat patches $N_p$ (ecosystem size) and branching probability $P_b$ (ecosystem complexity). Each habitat patch is assigned to be a branching patch (including upstream terminals) with probability $P_b$ or a non-branching patch with probability $1-P_b$. In this framework, an individual branch is a consecutive series of non-branching patches terminated at a branching patch; therefore, the number of habitat patches in branch $q$, $n_{p,q}$, is a realization of a random variable drawn from a geometric distribution $n_{p,q} \sim Ge(P_{b})$[@teruiMetapopulationStabilityBranching2018] but conditional on $\sum_{q=1}^{N_b}n_{p,q} = N_p$ ($N_b$ is the number of branches in a network). The number of branches $N_b$ is identical to the number of branching habitat patches in a network, or the number of 'success' out of $N_p$ trials. This leads to an assumption that $N_b$ is a realization of a random variable drawn from a binomial distribution as $N_b \sim Binomial(N_p, P_b)$. This framework has two merits. First, it reflects observed patterns of branch length distribution, which is known to follow an exponential distribution (a continuous version of a geometric distribution)[@rodriguez-iturbeFractalRiverBasins2001]. Second, it preserves the fractal nature of branching patterns[@peckhamReformulationHortonLaws1999].

@@

$\Omega_{xy}$ denotes the temporal covariance of environmental conditions between patch $x$ and $y$, which is assumed to decay exponentially with increasing distance between the patches $d_{xy}$. The parameter $\phi$ determines the degree of the distance decay of environmental correlates.

@@

$$
\xi_{ix}(t) = \frac{\sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}{\sum_{x} \sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}
$$
where $d_{xy}$ is the separation distance between patch $x$ and $y$. The parameter $\theta$ regulates the dispersal distance of a species that follows an exponential distribution ($\theta^{-1}$ corresponds to the expected dispersal distance). The expected number of immigrants is calculated as $I_{ix}(t) = \xi_{ix}(t)\sum_x^{N_p} E_{ix}$.

### Simulation
Based on the results of extensive sensitivity analysis (**Supplementary Information**), we set four landscape and eight ecological scenarios, resulting in a total of `r n_scenario` sets of parameter combinations. For landscape scenarios, we changed values of environmental variation at headwaters ($\sigma_h$) and the degree of local environmental noise ($\sigma_l$) separately to create contrasting patterns of habitat heterogeneity. Specifically, we used two values (0.01 and 1) for each of the SD parameters. When $\sigma_h > \sigma_l$, branching exhibits important effects on habitat heterogeneity in the ecosystem because headwaters are the primary source of environmental variation. When $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity in the ecosystem.

For ecological scenarios, we varied three ecological parameters that are relevant for dispersal and interspecific competition: dispersal distance ($\theta =$ `r unique(dat_sim$theta)`), dispersal probability ($p_d =$ `r unique(dat_sim$p_dispersal)`), and the maximum value of competition coefficients ($\alpha_{max} =$ `r unique(dat_sim$max_alpha)`). This setup results in eight ecological scenarios with different levels of dispersal and competition.

We allowed interspecific variation in niche optimum and width, which were drawn from uniform distributions as $\mu_i \sim Unif(-1, 1)$ and $\sigma_{niche,i} \sim Unif(0.1, 1)$. We used fixed values for the following parameters: maximum reproductive rate ($r_{0,i} =$ 4), niche cost ($\nu =$ `r unique(dat_sim$niche_cost)`), the degree of temporal fluctuation in abiotic environments ($\sigma_z =$ `r unique(dat_sim$sd_env)`), and the degree of spatial autocorrelation in temporal environmental dynamics ($\phi =$ `r unique(dat_sim$phi)`).

Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 200 time steps for initialization and 200 time steps for burn-in) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`). During the first 200 time steps (an initialization period), we initialized the simulation by seeding each habitat patch with populations of each species drawn from a Poisson distribution with a mean of 0.5. We repeated the seeding every 10 time steps during the initialization to allow species to establish their populations if they can grow from low abundances. After the initialization period, we ran 200 time steps as a burn-in period with no seeding to minimize influences of initial conditions. We saved the last 1000 time steps to estimate the temporal averages of $\alpha$, $\beta$, and $\gamma$ diversity. We removed simulation replicates in which no species established populations over the initial 400 time steps (`r sum(dat_sim$alpha_div == 0)` replicates out of `r nrow(dat_sim)` replicates). Ecosystem size and complexity were drawn randomly from uniform distributions as: $N_p \sim Unif(10, 150)$ and $P_b \sim Unif(0.01, 0.99)$ ($N_p$ was rounded to the nearest integer before running simulations).

@@

$$
\mu_i = \xi_0 + \xi_1~log_{10}~WA_i + \xi_2~log_{10}~P_{b,i} + \xi_3~region_i + \sum_k \xi_k x_{k,i}
$$
where $\xi_k$ are the intercept and regression coefficients for linear predictors. Meanwhile, the region-specific model assumes the effects of watershed area and branching probability differ between regions (i.e., interactions between region and watershed area/branching probability).

@@

$$
Bayes~factor = exp(\Delta BIC_{10}/2)
$$
where $\Delta BIC_{10} = BIC(M1) - BIC(M0)$.

>>>>

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  html_notebook: default
  word_document: default
abstract:
  The species-area relationship (the SAR), a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and has formed the basis of area-based conservation strategies. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity), and its role in scaling biodiversity is largely unknown. Here, we hypothesized that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity (area x habitat heterogeneity), two factors that regulate watershed-scale species richness. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two biogeographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the pervasive conflict between biodiversity conservation and societal demands.

@@

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@rinaldoEvolutionSelectionRiver2014; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grantLivingBranchesPopulation2007; @rinaldoRiverNetworksEcological2018;@tonkinRoleDispersalRiver2018]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure[@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @teruiMetapopulationStabilityBranching2018], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@altermattDiversityRiverineMetacommunities2013; @carraraComplexInteractionDendritic2014; @carraraDendriticConnectivityControls2012; @teruiMetapopulationStabilityBranching2018]. However, most research has explored the consequences of branching complexity for local community structure[@altermattRiverNetworkProperties2013; @brownDendriticNetworkStructure2010] or has relied solely on theoretical arguments with limited replications of branching architecture[@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services[@dudgeonFreshwaterBiodiversityImportance2006].

@@

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches which local communities inhabit[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a] (**Figure 1**). Two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see **Methods** for model details).

Given the results of extensive sensitivity analysis (**Supplementary Information**), we considered `r n_scenario` simulation scenarios comprising a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries [@chezikRiverNetworksDampen2017; @mooreEmergentStabilityLarge2015; @petersonModellingDendriticEcological2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities[@mooreSynchronizationPortfolioPerformance2010; @poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see **Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Information**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

Our theoretical analysis yielded results consistent with our prediction. Ecosystem size and complexity both increased $\gamma$ diversity under a realistic landscape scenario (**Figures 2 and 3**), where the environmental variation at headwaters ($\sigma_h$) is greater than the degree of local environmental noise ($\sigma_l$). The relationships had a characteristic of power-law (i.e., linear in a log-log scale) and were consistent under various ecological scenarios. The strength of competition and dispersal processes (dispersal distance and probability) did not change the relationships between $\gamma$ diversity and ecosystem properties (compare panels in **Figures 2 and 3**). Hence, the ecosystem size and complexity are expected to increase $\gamma$ diversity regardless of ecological scenarios.

@@

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data across two geographic regions: the Hokkaido island in Japan and the midwestern US (Midwest). These regions are located in comparable latitude ranges but support distinct fish communities (**Figure 1**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Methods** and **Supplementary Information**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area), or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). We combined this data set with geospatial information (watershed area, branching probability, annual mean temperature, annual cumulative precipitation, the fraction of forest, and dam density) to examine potential influences of macro-scale drivers of species diversity. Using this data set, we developed global and region-specific models for each of the diversity measures ($\alpha$, $\beta$, and $\gamma$ diversity) to examine whether observed patterns are consistent across the two geographic regions. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity by including interaction terms with the dummy binary variable of the region (Hokkaido = 0; Midwest = 1). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (see **Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

@@

The emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation[@poffHomogenizationRegionalRiver2007], landscape modification[@finlayStreamSizeHuman2011; @moritaEffectsHabitatFragmentation2002], and stream burial[@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research[@koningNetworkGrassrootsReserves2020; @stollSmallImpoverishedRegional2013; @sundermannRiverRestorationSuccess2011]. While large reserves or restoration sites are undoubtedly important[@koningNetworkGrassrootsReserves2020], our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin[@koningNetworkGrassrootsReserves2020].

While the prevailing evidence supports the importance of ecosystem size in scaling species diversity, ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial[@turnerLandscapeEcologyTheory2015] and aquatic ecosystems[@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. The dual dimensional approach - size and complexity - broadens viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the pervasive conflict between biodiversity conservation and societal demands. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

We depicted branching ecosystems as a bifurcating network of connected habitat patches[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a] (**Figure 1**). Habitat patches, or local communities, can be either non-branching or branching river sections with a unit length $l$, which defines the scale of local species interactions. Two parameters determine the geometric properties of simulated branching networks: the number of habitat patches $N_p$ (ecosystem size) and branching probability $P_b$ (ecosystem complexity). Each habitat patch is assigned to be a branching patch (including upstream terminals) with probability $P_b$ or a non-branching patch with probability $1-P_b$. In this framework, an individual branch is a consecutive series of non-branching patches terminated at a branching patch; therefore, the number of habitat patches in branch $q$, $n_{p,q}$, is a realization of a random variable drawn from a geometric distribution $n_{p,q} \sim Ge(P_{b})$[@teruiMetapopulationStabilityBranching2018] but conditional on $\sum_{q=1}^{N_b}n_{p,q} = N_p$ ($N_b$ is the number of branches in a network). The number of branches $N_b$ is identical to the number of branching habitat patches in a network, or the number of 'success' out of $N_p$ trials. This leads to an assumption that $N_b$ is a realization of a random variable drawn from a binomial distribution as $N_b \sim Binomial(N_p, P_b)$. This framework has two merits. First, it reflects observed patterns of branch length distribution, which is known to follow an exponential distribution (a continuous version of a geometric distribution)[@rodriguez-iturbeFractalRiverBasins2001]. Second, it preserves the fractal nature of branching patterns[@peckhamReformulationHortonLaws1999].

@@

$\Omega_{xy}$ denotes the temporal covariance of environmental conditions between patch $x$ and $y$, which is assumed to decay exponentially with increasing distance between the patches $d_{xy}$ (the shortest network path [measured as the number of habitat patches] to reach a target patch). The parameter $\phi$ determines the degree of the distance decay of environmental correlates.

@@

$$
\xi_{ix}(t) = \frac{\sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}{\sum_{x} \sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}
$$
where $d_{xy}$ is the separation distance between patch $x$ and $y$ (the shortest network path [measured as the number of habitat patches] to reach a target patch). The parameter $\theta$ regulates the dispersal distance of a species that follows an exponential distribution ($\theta^{-1}$ corresponds to the expected dispersal distance). The expected number of immigrants is calculated as $I_{ix}(t) = \xi_{ix}(t)\sum_x^{N_p} E_{ix}$.

### Simulation
We used `r n_scenario` sets of parameter combinations (i.e., simulation scenarios) based on the results of extensive sensitivity analysis (**Supplementary Information**). The parameter combinations comprises four landscape and eight ecological scenarios. Landscape scenarios distinguish spatial patterns of ecosystem's habitat heterogeneity, while ecological scenarios change ecological traits of simulated species. For landscape scenarios, we used four combinations of environmental variation at headwaters ($\sigma_h = 0.01, 1$) and the degree of local environmental noise ($\sigma_l = 0.01, 1$). The inequality of $\sigma_h$ and $\sigma_l$ controls the coupling of branching complexity and ecosystem's habitat heterogeneity. For ecological scenarios, we varied three ecological parameters that are relevant for dispersal and interspecific competition: dispersal distance ($\theta =$ `r unique(dat_sim$theta)`), dispersal probability ($p_d =$ `r unique(dat_sim$p_dispersal)`), and the maximum value of competition coefficients ($\alpha_{max} =$ `r unique(dat_sim$max_alpha)`). This setup results in eight ecological scenarios with different levels of dispersal and competition.

We allowed interspecific variation in niche optimum and width, which were drawn from uniform distributions as $\mu_i \sim Unif(-1, 1)$ and $\sigma_{niche,i} \sim Unif(0.1, 1)$. We used fixed values for the following parameters: maximum reproductive rate ($r_{0,i} =$ 4), niche cost ($\nu =$ `r unique(dat_sim$niche_cost)`), the degree of temporal environmental variation ($\sigma_z =$ `r unique(dat_sim$sd_env)`), and the extent of spatial autocorrelation in temporal environmental dynamics ($\phi =$ `r unique(dat_sim$phi)`).

Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 200 time steps for initialization and 200 time steps for burn-in) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`). During the first 200 time steps (an initialization period), we initialized the simulation by seeding each habitat patch with populations of each species drawn from a Poisson distribution with a mean of 0.5. We repeated the seeding every 10 time steps during the initialization to allow species to establish their populations if they can grow from low abundances. After the initialization period, we ran 200 time steps as a burn-in period with no seeding to minimize influences of initial conditions. We saved the last 1000 time steps to estimate the temporal averages of $\alpha$, $\beta$, and $\gamma$ diversity. We removed simulation replicates in which no species established populations over the initial 400 time steps (`r sum(dat_sim$alpha_div == 0)` replicates out of `r nrow(dat_sim)` replicates). Ecosystem size and complexity were drawn randomly from uniform distributions as: $N_p \sim Unif(10, 150)$ and $P_b \sim Unif(0.01, 0.99)$ ($N_p$ was rounded to the nearest integer before running simulations). Values of simulation parameters were summarized in **Table S2**. We performed simulations using R version 4.0.2[@rcoreteamLanguageEnvironmentStatistical2020].

@@

$$
\mu_i = \xi_0 + \xi_1~log_{10}~WA_i + \xi_2~log_{10}~P_{b,i} + \xi_3~region_i + \sum_k \xi_k x_{k,i}
$$

where $\xi_k$ are the intercept and regression coefficients for linear predictors. Meanwhile, the region-specific model assumes the effects of watershed area and branching probability differ between regions (i.e., interactions between region and watershed area/branching probability).

@@

$$
Bayes~factor = exp(\Delta BIC_{10}/2)
$$
